Rice yield improvements by plant breeding are offset by decreases over time

### 2017-05-04 output/breed_paper.Rnw
Adjust to only show RES

\documentclass[12pt,]{elsarticle}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile, textcomp}
\usepackage{setspace}
\usepackage{etoolbox}
\doublespacing
\usepackage{lineno}
\usepackage{lscape} %% Harvard
\bibliographystyle{model2-names}\biboptions{authoryear}
\pagenumbering{gobble}
%% For figure numbers in supplementary info
%% from http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
\newcommand{\beginsupplement}{%
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
}
\usepackage{Sweave}
\SweaveOpts{prefix.string=graphics/report}

@@

\title{Rice yield improvements by plant breeding are offset by
  decreases over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}

@@

@@

\subsection{Site description}

The Rice Experiment Station (RES) of the California Cooperative Rice
Research Foundation

@@

Linear model of yield of a cultivar as a response to the number of years since it
was entered into the trials. Corrections for year, site, year:site,
and cultivar differences are made according to:

\begin{equation}
 yield = \alpha + \beta_{years\ in\ trial}* x_{years\ in\ trials} + \beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}
\end{equation}

Only medium-grained, released cultivars were included in the analysis
and data from the San Joaquin trials, which are not representative of
typical CA rice production (frequent exposure to cold, drill-seeded,
etc.) were excluded. This reduced the data from \Sexpr{nrow(all_vt_result)} to
\Sexpr{nrow(mod_data)} observations. Included cultivars are:
\Sexpr{paste(gsub('^M', 'M-', unique(mod_data[,'id'])), collapse = ', ')}. The
official release years were used for each cultivar for the year of
release (i.e., not the year seed was first commercially available).

A similar but separate model was used to estimate whether newer
releases had a yield advantage over older ones:

\begin{equation}
yield = \alpha + \beta_{release\ year}*x_{release\ year} +
\beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}
\end{equation}

Again, the estimate was conditioned on differences between cultivars,
sites, years, and site:years.

@@

<<fig=TRUE, echo=FALSE>>=
yrs <- 1980:2015
pp <- posterior_predict(release_fit, re.form = NA, newdata = data.frame(release_yr_c = yrs - 2000))
pp <-  (pp * 1000 + 9000) * 1.12

@@

res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_lb[res] * 1.12,
                           mod_data$id[res])

@@

plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8000, 13000),
     xlim = c(1980,2019))
abline(a = (((fixef(RES_rel_fit)[1] * 1000) + 9000) * 1.12) -
           (2000 * fixef(RES_rel_fit)[2] * 1000),
       b = fixef(RES_rel_fit)[2] * 1000, col = "grey50")
with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr, y = plot_data$yield_kg,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.8)

text(1990, 8000, bquote(.(round(fixef(release_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))

@@

@@

% <<fig=TRUE, echo=FALSE>>=
% yrs_in <- -6:34
% pp <- posterior_predict(yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
% pp <-  (pp * 1000 + 9000) * 1.12

@@

% plot_data = as.data.frame(tapply(mod_data$yield_lb * 1.12, mod_data$yrs_in_trial, mean))
% plot_data$se = tapply(mod_data$yield_lb * 1.12, mod_data$yrs_in_trial, function(x)
%     sd(x)/sqrt(length(x)))
% plot_data$tmp = as.numeric(row.names(plot_data))
% colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")

% plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
%      xlab = "Year since entry",
%      ylab = expression("Yield ( kg ha"^-1~")"),
%      ylim = c(8700, 11200))
% with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
%                          y1 = yield_kg - yield_se))
% abline(a = ((fixef(yrs_fit)[1] * 1000) + 9000) * 1.12, b = fixef(yrs_fit)[2] * 1000)
% # lines(x = yrs_in, y = ints[2,], lty = 2) 
% # polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
% text(30, 4500, bquote(.(round(fixef(yrs_fit)[2] * 1000, 1))~lbs~year^-1))

% @ 
% \caption{There is a small decrease in yield over the time since a
  % cultivar was entered into the trial using the full data
  % set. However, the effect is small (\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)}
  % lbs year\textsuperscript{-1}) in relation to the variability
  % between sites and years.}
% \label{fig:yrs}
% \end{figure}

<<echo=FALSE>>=
all_CI <- round(posterior_interval(yrs_fit, pars = "yrs_in_trial") * 1000, 1)
RES_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * 1000, 1)
@

As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
over time once a cultivar is released, though the estimated effect is
small compared to the variability in the data set. This decrease is estimated to be
\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)} lbs
year\textsuperscript{-1}. The oldest cultivar in the trials has lost
an estimated \Sexpr{round(fixef(yrs_fit)[2] * 1000 * 36 * 1.12, 1)} kg ha\textsuperscript{-1} since release. Using just data from the Rice
Experiment Station (RES), the estimated effect is more drastic
(an estimated \Sexpr{round(fixef(RES_yrs_fit)[2] * 1000 * 1.12, 1)} kg
ha\textsuperscript{-1} year\textsuperscript{-1}), but with a wider credible interval
(\Sexpr{RES_CI[1]} -- \Sexpr{RES_CI[2]} vs \Sexpr{all_CI[1]} --
\Sexpr{all_CI[2]}).

<<fig=TRUE, echo=FALSE>>=
yrs_in <- -7:35
pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
pp <-  (pp * 1000 + 9000) * 1.12

RES = res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_lb[res] * 1.12,
                           mod_data$yrs_in_trial[res])

@@

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since entry",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(7000, 13000))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_yrs_fit)[1] * 1000) + 9000) * 1.12,
       b = fixef(RES_yrs_fit)[2] * 1000, col = "grey50")
ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
text(5, 7000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))

@
\caption{The relationship between yield and the years since entry for
  medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. RES is considered
  the best managed site, and has the most plots per year of any of the
  sites. The estimated decrease over time is estimated to be stronger
  using only these data (\Sexpr{round(fixef(RES_yrs_fit)[2] * 1000, 1)}
  lbs year\textsuperscript{-1}), primarly due to lower variability.}
\label{fig:RES_year}
\end{figure}

plot_data = lapply(tars, function(x) {
    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
    tmp = prep_plot_data(dd$yield_lb * 1.12, dd$year)
    tmp$year = as.numeric(row.names(tmp))
    tmp
    })

invisible(sapply(1:2, function(i){
    plot(mean ~ year,
         data = plot_data[[i]],
         xlim = c(1995, 2016), ylim = c(6500,14500),
         pch = 16, main = gsub("M","M-", tars[i]),
         ylab = "",
         yaxt = "n",
         xlab = "Years after release")
    segments(x0 = plot_data[[i]]$year,
             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
    if(i == 1){
        axis(2)
        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }
#     abline(lm(yield_lb*1.12 ~ year,
    #               data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
    # lty = 2)

    abline(h = mean(mod_data$yield_lb[(mod_data$id == tars[i] &
                                       mod_data$site == "RES")] * 1.12),
           lty = 1, col = "grey50")

# invisible(sapply(tars, function(x){
#     plot(yield_lb * 1.12 ~ year,
#          data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
#          xlim = c(1995, 2016), ylim = c(6500,14500),
#          pch = 16,
#          col = rgb(0,0,0,0.5),
#          main = gsub("M", "M-", x),
#          ylab = "",
#          yaxt = "n",
#          xlab = "Year")
#     if(x == "M202"){
#         axis(2)
#         mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
#     }
#     # abline(lm(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
#     abline(h = mean(mod_data$yield_lb[(mod_data$id == x & mod_data$site == "RES")] * 1.12),
#            lty = 2)
# }))

@
\caption{The performance of two major cultivars at the CA Rice
  Experiment Station (RES) over the period 1995 -- 2015. The dashed line is the average
  yield for each cultivar over the period which it was planted at the
  RES. The newer cultivar, M-209, appears to have a slight yield
  advantage over the older cultivar, M-202.}
\label{fig:res_yrs}
\end{figure}

@@

<<fig=TRUE, echo=FALSE, width = 8, height = 5>>=
par(mfrow = c(1, length(tars)),
    mar = c(4,0,2,0),
    oma = c(4,4,0.5,0.5))

plot_data = lapply(tars, function(x) {
    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
    tmp = prep_plot_data(dd$yield_lb * 1.12, dd$yrs_in_trial)
    tmp$yrs_in_trial = as.numeric(row.names(tmp))
    tmp
    })

invisible(sapply(1:2, function(i){
    plot(mean ~ yrs_in_trial,
         data = plot_data[[i]],
         xlim = c(-6, 36), ylim = c(6500,14500),
         pch = 16, main = gsub("M","M-", tars[i]),
         ylab = "",
         yaxt = "n",
         xlab = "Years after release")
    segments(x0 = plot_data[[i]]$yrs_in_trial,
             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
    if(i == 1){
        axis(2)
        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }
    abline(lm(yield_lb*1.12 ~ yrs_in_trial,
              data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
           lty = 1, col = "grey50")
}))

@
\caption{The performance of two major cultivars at the CA Rice
  Experiment Station (RES) since the year each cultivar was
  released. Both M-202 and M-206 have similar intercepts, suggesting
  similar yield at the time of release and an apparent yield advantage
  of the newer cultivar, M-209, might be explained by a yield loss of
  M-202 over time.}
\label{fig:res_since_release}
\end{figure}

Contrary to the estimates for decreases once a cultivar is released, there was a
small increase over time (Fig. \ref{fig:release}), though it was small
(\Sexpr{round(fixef(release_fit)[2] * 1000, 1)} lbs
year\textsuperscript{-1}). This suggests that the overall progress on
yield over the period from 1980 -- 2016 is approximately
\Sexpr{round(fixef(release_fit)[2] * 1000 * 36, 1)} lbs acre\textsuperscript{-1}.

@@

% <<fig=TRUE, echo=FALSE>>=
% yrs <- 1980:2015
% pp <- posterior_predict(RES_rel_fit, re.form = NA,
%                         newdata = data.frame(release_yr_c = yrs - 2000))
% pp <-  pp * 1000 + 9000

@@

% plot(yield_lb ~ release_yr, data = mod_data[RES,], pch = 16, col = rgb(0,0,0,0.25),
%      xlab = "Year of Release",
%      ylab =  expression("Yield ( lbs acre"^-1~")"))
% lines(x = yrs, y = ints[2,], lty = 2) 
% polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
% text(2010, 4500, bquote(.(round(fixef(RES_rel_fit)[2] * 1000, 1))~lbs~year^-1))
% @

@@

There is evidence of a small decrease in yield over time once a cultivar is
released. This complicates comparisons between new entries and checks,
since the check will decrease over time. This decrease will manifest
as an apparent larger yield advantage of the new entry over the check.
There is evidence of a small increase in yield with each new released
cultivar. Therefore, it appears that successive cultivar releases are
not merely recovering to the old benchmark but are surpassing
it. However, progress is slower than previously estimated
(\Sexpr{round(fixef(release_fit)[2] * 1000, 1)} lbs
acre\textsuperscript{-1} year\textsuperscript{-1} increase verses a
previous estimate of 54 lbs acre\textsuperscript{-1}
year\textsuperscript{-1}).

@@

<<label=plot1, fig=TRUE, echo=FALSE>>=
boxplot(yield_lb ~ year, data = mod_data,
        xlab = "Year", ylab = expression("Yield ( lb ac"^-1~")"))
@
  \caption{Distribution of yield over the period 1995 -- 2015.} \label{fig:plot}
\end{figure}

@@

  boxplot(yield_lb ~ id, data = mod_data,
          xlab = "Cultivar",
          ylab = "Yield (lbs/ac)")
@
\caption{Boxplot of yields for released medium grain cultivars for
  the period 1995--2015.}
\label{fig:raw_yield}
\end{figure}

>>>>

\documentclass[12pt,]{elsarticle}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile, textcomp}
\usepackage{setspace}
\usepackage{etoolbox}
\doublespacing
\usepackage{lineno}
\usepackage{lscape} %% Harvard
\bibliographystyle{model2-names}\biboptions{authoryear}
\pagenumbering{gobble}
%% For figure numbers in supplementary info
%% from http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
\newcommand{\beginsupplement}{%
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
}
\usepackage{Sweave}
\SweaveOpts{prefix.string=graphics/report}

@@

\title{Rice yield improvements by plant breeding are offset by
  decreases over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\author[ucd]{David MacKill}
\author[ucd]{Jim E. Hill}
\author[res]{Kent McKenzie}
\author[ucd]{Bruce A. Linquist}

\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}

@@

\address[res]{Rice Experiment Station, California Cooperative Rice
Research Foundation, Biggs, CA 95917}

@@

\subsection{Site and data description}

The Rice Experiment Station (RES) of the California Cooperative Rice
Research Foundation (CCRRF) is located northwest of Biggs, CA (39.4648,
-121.7342), and has been the central location for the development of
improved rice varieties suited for CA since 1969. The climate at the
RES is Mediterranian, characterized by mild winters during which most
of the annual precipitation occurs (\Sexpr{18*24.5} -- \Sexpr{25*24.5}
mm annually) and warm summers largely free of precipitation
events. Soils at the RES are classified as Esquon-Neerdobe clays with
roughly 2\% SOM \citep{ssurgo} in the top 15cm. As the central
location for the breeding program, the RES is researcher-managed (as
opposed to farmer-managed trials) and conducts the bulk of the trials
for the Statewide Variety Trials (approximately 30\% of the total
plots are planted at RES). For these reasons, we focused our
investigation on the RES to avoid complications due to site, climate, and
management differences at the other 6 -- 8 farmer-managed trials.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against named and released cultivars (checks) at
the RES in multiple trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as medium grain cultivars constitute roughly
90\% of the planted area in CA. This lead to a final data set of
\Sexpr{format(nrow(mod_data), big.mark = ',')} observations with
\Sexpr{length(unique(mod_data[,'id']))} cultivars represented,
constituting every publicly released medium grain variety developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of release (typically 3 -- 5 years after introduction to the
trials but prior to widespread commercial seed availability) and the
time elapse (in years) since release was calculated.

@@

Yield was modeled as a response to the number of years since the
variety was officially released using a mixed-effects linear
model. Differences between years and cultivars were included in the
model as random effects according to:

\begin{equation}
 yield_i = \alpha + \beta_{years\ since\ release\ i}* x_{years\ since\
   release\ i} + \beta_{year\ i} + \beta_{cultivar\ i}
\end{equation}

Where $yield_i$ is the grain yield for plot $i$, $\alpha$ is the
average yield for the RES across varieties and years (intercept),
$\beta_{years\ since\ release\ i}$ is effect of $x_{years\ since\
release\ i}$, the number of years $i$'s cultivar has been in the
trials (negative for years prior to release, zero at the year of
release), $\beta_{year}$ is the random effect of year, and
$\beta_{cultivar}$ is the random effect of cultivar. A similar but
separate model was used to estimate the yield advantage of newer
releases over older ones:

\begin{equation}
yield_i = \alpha + \beta_{release\ year\ i}*x_{release\ year\ i} +
\beta_{year\ i} + \beta_{cultivar\ i}
\end{equation}

Where $\beta_{release\ year\ i}$ is effect of $x_{release\ year\ i}$,
the year of release for $i$'s cultivar. All other parameters are
similar to Eq. 1.

Data was processed and models were fit in R, an environment for
statistical computing \citep{R}. Models were fit using the 'rstanarm'
package \citep{rstanarm}, an interface to Stan, a language for
probabilistic programming \citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results.

@@

<<echo=FALSE>>=
rel_CI = round(posterior_interval(RES_rel_fit, pars = "release_yr_c") * scl, 1)

posterior_interval(RES_rel_fit, pars = "release_yr_c") * scl
m202_v_206 = ((posterior_predict(RES_rel_fit, newdata = data.frame(release_yr_c = c(-15, 4)),
                                 re.form = NA) * scl) + center)
m202_v_206 = m202_v_206[,2] - m202_v_206[,1]
m202_v_206 = round(quantile(m202_v_206, p = c(0.05, 0.5, 0.95)), 1)
@

There was a positive relationship between the year of release and
yield, with an estimated \Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)}
kg ha\textsuperscript{-1} year\textsuperscript{-1} increase (90\% CI:
\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]})
(Fig. \ref{fig:release}). Comparing the two dominate cultivars M-202,
which was widely planted from the late 1980s until the early 2000s,
and M-206, currently the most widely planted CRRFS cultivar
(approximately 50\% of planted area), there is an estimated
\Sexpr{m202_v_206[2]} kg ha\textsuperscript{-1} yield advantage of M-206
over M-202 (Figs \ref{fig:release} \& \ref{fig:compare_yrs}).

\begin{figure}[ht]

<<fig=TRUE, echo=FALSE>>=
yrs <- 1980:2015
pp <- posterior_predict(RES_rel_fit, re.form = NA, newdata = data.frame(release_yr_c = yrs - 2000))
pp <-  (pp * scl + center)

@@

res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_kg[res] ,
                           mod_data$id[res])

@@

plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8000, 12500),
     xlim = c(1980,2019))
abline(a = (((fixef(RES_rel_fit)[1] * scl) + center) ) -
           (2000 * fixef(RES_rel_fit)[2] * scl),
       b = fixef(RES_rel_fit)[2] * scl, col = "grey50",
       lty = 2)
with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr, y = plot_data$yield_kg,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.8)

text(1990, 8000, bquote(.(round(fixef(RES_rel_fit)[2] * scl , 1))~kg~ha^-1~year^-1))

@@

plot_data = lapply(tars, function(x) {
    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
    tmp = prep_plot_data(dd$yield_kg , dd$year)
    tmp$year = as.numeric(row.names(tmp))
    tmp
    })

invisible(sapply(1:2, function(i){
    plot(mean ~ year,
         data = plot_data[[i]],
         xlim = c(1995, 2016), ylim = c(7500,13500),
         pch = 16, main = gsub("M","M-", tars[i]),
         ylab = "",
         yaxt = "n",
         xlab = "Years after release")
    segments(x0 = plot_data[[i]]$year,
             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
    if(i == 1){
        axis(2)
        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }
#     abline(lm(yield_kg*1.12 ~ year,
    #               data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
    # lty = 2)

    abline(h = mean(mod_data$yield_kg[(mod_data$id == tars[i] &
                                       mod_data$site == "RES")] ),
           lty = 1, col = "grey50")

# invisible(sapply(tars, function(x){
#     plot(yield_kg  ~ year,
#          data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
#          xlim = c(1995, 2016), ylim = c(6500,14500),
#          pch = 16,
#          col = rgb(0,0,0,0.5),
#          main = gsub("M", "M-", x),
#          ylab = "",
#          yaxt = "n",
#          xlab = "Year")
#     if(x == "M202"){
#         axis(2)
#         mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
#     }
#     # abline(lm(yield_kg ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
#     abline(h = mean(mod_data$yield_kg[(mod_data$id == x & mod_data$site == "RES")] ),
#            lty = 2)
# }))

@
\caption{The performance of two major cultivars at the CA Rice
  Experiment Station (RES) over the period 1995 -- 2015. The line is the average
  yield for each cultivar over the period which it was planted at the
  RES. The newer cultivar, M-206, appears to have a slight yield
  advantage over the older cultivar, M-202.}
\label{fig:compare_yrs}
\end{figure}

@@

% <<fig=TRUE, echo=FALSE>>=
% yrs_in <- -6:34
% pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
% pp <-  (pp * scl + center)

@@

% plot_data = as.data.frame(tapply(mod_data$yield_kg , mod_data$yrs_in_trial, mean))
% plot_data$se = tapply(mod_data$yield_kg , mod_data$yrs_in_trial, function(x)
%     sd(x)/sqrt(length(x)))
% plot_data$tmp = as.numeric(row.names(plot_data))
% colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")

% plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
%      xlab = "Year since entry",
%      ylab = expression("Yield ( kg ha"^-1~")"),
%      ylim = c(8700, 11200))
% with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
%                          y1 = yield_kg - yield_se))
% abline(a = ((fixef(RES_yrs_fit)[1] * scl) + center) , b = fixef(RES_yrs_fit)[2] * scl)
% # lines(x = yrs_in, y = ints[2,], lty = 2) 
% # polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
% text(30, 4500, bquote(.(round(fixef(RES_yrs_fit)[2] * scl, 1))~lbs~year^-1))

% @ 
% \caption{There is a small decrease in yield over the time since a
  % cultivar was entered into the trial using the full data
  % set. However, the effect is small (\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)}
  % lbs year\textsuperscript{-1}) in relation to the variability
  % between sites and years.}
% \label{fig:yrs}
% \end{figure}

<<echo=FALSE>>=
all_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * scl, 1)
RES_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * scl, 1)
@

As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
over time once a cultivar is released, though the estimated effect is
small compared to the variability in the data set. This decrease is estimated to be
\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)} lbs
year\textsuperscript{-1}. The oldest cultivar in the trials has lost
an estimated \Sexpr{round(fixef(RES_yrs_fit)[2] * scl * 36 , 1)} kg ha\textsuperscript{-1} since release. Using just data from the Rice
Experiment Station (RES), the estimated effect is more drastic
(an estimated \Sexpr{round(fixef(RES_yrs_fit)[2] * scl , 1)} kg
ha\textsuperscript{-1} year\textsuperscript{-1}), but with a wider credible interval
(\Sexpr{RES_CI[1]} -- \Sexpr{RES_CI[2]} vs \Sexpr{all_CI[1]} --
\Sexpr{all_CI[2]}).

\begin{figure}[ht]

<<fig=TRUE, echo=FALSE>>=
yrs_in <- -7:35
pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
pp <-  (pp * scl + center)

RES = res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_kg[res] ,
                           mod_data$yrs_in_trial[res])

@@

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since release",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8000, 12500))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_yrs_fit)[1] * scl) + center) ,
       b = fixef(RES_yrs_fit)[2] * scl, col = "grey50",
       lty = 2)
ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
text(5, 8000, bquote(.(round(fixef(RES_yrs_fit)[2] * scl , 1))~kg~ha^-1~year^-1))

@
\caption{The relationship between yield and the years since release for
  medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. RES is considered
  the best managed site, and has the most plots per year of any of the
  sites. The estimated decrease over time is estimated to be stronger
  using only these data (\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)}
  lbs year\textsuperscript{-1}), primarly due to lower variability.}
\label{fig:RES_year}
\end{figure}

@@

<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
# par(mfrow = c(1, length(tars)),
#     mar = c(4,0,2,0),
#     oma = c(4,4,0.5,0.5))

plot_data = lapply(tars, function(x) {
    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
    tmp = prep_plot_data(dd$yield_kg , dd$yrs_in_trial)
    tmp$yrs_in_trial = as.numeric(row.names(tmp))
    tmp
    })

# invisible(sapply(1:2, function(i){
#     plot(mean ~ yrs_in_trial,
#          data = plot_data[[i]],
#          xlim = c(-6, 36), ylim = c(6500,14500),
#          pch = 16, main = gsub("M","M-", tars[i]),
#          ylab = "",
#          yaxt = "n",
#          xlab = "Years after release")
#     segments(x0 = plot_data[[i]]$yrs_in_trial,
#              y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
#              y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
#     if(i == 1){
#         axis(2)
#         mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
#     }
#     abline(lm(yield_kg*1.12 ~ yrs_in_trial,
#               data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
#            lty = 1, col = "grey50")
# }))

plot(mean ~ yrs_in_trial,
     data = plot_data[[1]],
     xlim = c(-6, 32), ylim = c(6500,14500),
     pch = 16, main = gsub("M","M-", tars[i]),
     ylab = expression("Yield ( kg ha"^-1~")"),
     xlab = "Years after release")
segments(x0 = plot_data[[1]]$yrs_in_trial,
         y0 = plot_data[[1]]$mean + plot_data[[1]]$se,
         y1 = plot_data[[1]]$mean - plot_data[[1]]$se)
points(mean ~ yrs_in_trial,
       data = plot_data[[2]],
       pch = 17, col = "grey40")
segments(x0 = plot_data[[2]]$yrs_in_trial,
         y0 = plot_data[[2]]$mean + plot_data[[2]]$se,
         y1 = plot_data[[2]]$mean - plot_data[[2]]$se,
         col = "grey40")
abline(lm(yield_kg ~ yrs_in_trial,
          data = mod_data[(mod_data$id %in% tars & mod_data$site == "RES"),]),
       lty = 1, col = "grey50")
legend("topright", legend = gsub("M","M-", tars),
       pch = c(16, 17), col = c("black","grey40"))

@
\caption{The performance of two major cultivars at the CA Rice
  Experiment Station (RES) since the year each cultivar was
  released. Both M-202 ($\bullet$) and M-206 ($\blacktriangle$) have similar
  intercepts, suggesting similar yield at the time of release and an
  apparent yield advantage of the newer cultivar, M-206, might be
  explained by a yield loss of M-202 over time. Points represent the
  mean, and bars (|) the standard error.}
\label{fig:res_since_release}
\end{figure}

Contrary to the estimates for decreases once a cultivar is released, there was a
small increase over time (Fig. \ref{fig:release}), though it was small
(\Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)} lbs
year\textsuperscript{-1}). This suggests that the overall progress on
yield over the period from 1980 -- 2016 is approximately
\Sexpr{round(fixef(RES_rel_fit)[2] * scl * 36, 1)} lbs acre\textsuperscript{-1}.

@@

% <<fig=TRUE, echo=FALSE>>=
% yrs <- 1980:2015
% pp <- posterior_predict(RES_rel_fit, re.form = NA,
%                         newdata = data.frame(release_yr_c = yrs - 2000))
% pp <-  pp * scl + center

@@

% plot(yield_kg ~ release_yr, data = mod_data[RES,], pch = 16, col = rgb(0,0,0,0.25),
%      xlab = "Year of Release",
%      ylab =  expression("Yield ( lbs acre"^-1~")"))
% lines(x = yrs, y = ints[2,], lty = 2) 
% polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
% text(2010, 4500, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~lbs~year^-1))
% @

@@

There is evidence of a small decrease in yield over time once a cultivar is
released. This complicates comparisons between new entries and checks,
since the check will decrease over time. This decrease will manifest
as an apparent larger yield advantage of the new entry over the check.
There is evidence of a small increase in yield with each new released
cultivar. Therefore, it appears that successive cultivar releases are
not merely recovering to the old benchmark but are surpassing
it. However, progress is slower than previously estimated
(\Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)} lbs
acre\textsuperscript{-1} year\textsuperscript{-1} increase verses a
previous estimate of 54 lbs acre\textsuperscript{-1}
year\textsuperscript{-1}).

@@

<<label=plot1, fig=TRUE, echo=FALSE>>=
boxplot(yield_kg ~ year, data = mod_data,
        xlab = "Year", ylab = expression("Yield ( lb ac"^-1~")"))
@
  \caption{Distribution of yield over the period 1995 -- 2015.} \label{fig:plot}
\end{figure}

@@

  boxplot(yield_kg ~ id, data = mod_data,
          xlab = "Cultivar",
          ylab = "Yield (lbs/ac)")
@
\caption{Boxplot of yields for released medium grain cultivars for
  the period 1995--2015.}
\label{fig:raw_yield}
\end{figure}
